Are translation equivalents special? Evidence from simulations and empirical data from bilingual infants

### 2021-12-16 paper/paper_TE_bilingual_vocabulary_model.tex
Updated paper

\abstract{
The acquisition of translation equivalents is often considered a special component of bilingual children's vocabulary development, as bilinguals have to learn words that share the same meaning across their two languages. This study examined three contrasting accounts for bilingual children's acquisition of translation equivalents relative to words that are first labels for a referent: the Avoidance Account whereby translation equivalents are harder to learn, the Preference Account whereby translation equivalents are easier to learn, and the Neutral Account whereby translation equivalents are similar to learn. To adjudicate between these accounts, Study 1 explored patterns of translation equivalent learning under a novel computational model --- the Bilingual Vocabulary Model --- which quantifies translation equivalent knowledge as a function of the probability of learning words in each language, and includes a bias parameter that varies the difficulty of learning translation equivalents according to each account. Study 2 tested model-derived predictions against vocabulary data from 200 French--English bilingual children aged 18--33 months. Results showed a close match between the model predictions and bilingual children's patterns of translation equivalent learning. At smaller vocabulary sizes, data matched the Preference Account, while at larger vocabulary sizes they matched the Neutral Account. Our findings show that patterns of translation equivalent learning emerge predictably from the word learning process, and reveal a qualitative shift in translation equivalent learning as bilingual children develop and learn more words.
}

@@

@@

\caption{Number of translation equivalents (TE) across different levels of vocabulary balance (BALANCE) in relation to word vocabulary (WORD; Panel A), dominant vocabulary size (DOM; Panel B), and non-dominant vocabulary size (NONDOM; Panel C). Row 1 represents the simulated data in Study 1 while holding the number of learnable words (LEARNABLE) constant at 600 and BIAS constant at 1. Row 2 represents the observed vocabulary data in Study 2.}\label{fig:fig1}
\end{figure}

@@

@@

Translation equivalent (TE) knowledge was first simulated across different developmental levels (as indicated by number of LEARNABLE words = 150, 300, 450, 600), in conjunction with a wide range of values for DOM and NONDOM. Again, BALANCE and word vocabulary (WORD) were calculated based on the values of DOM and NONDOM. The final simulated data set contained 166 data points (see Table 2 for a summary of the parameters used). Three scenarios of translation equivalent learning (TE) were then generated using the formula TE = BIAS × DOM×NONDOM/LEARNABLE. To illustrate the Avoidance Account, BIAS was set at .5 (i.e., TEs are 50\% less likely to be learned than singlets). To illustrate the Neutral Account, BIAS was set at 1 (i.e., TEs are equally learnable as singlets). Finally, to illustrate the Preference Account, BIAS was set at 1.5 (i.e., TE are 50\% more likely to be learned than singlets). In Figure 3, we illustrate the three different scenarios of simulated translation equivalent (TE) knowledge. Again, we continue to observe a pattern consistent with prediction 1a where, in all cases, simulated children with more balanced vocabularies (BALANCE) produced more translation equivalents (TE). Thus, overall relationships between BALANCE and TE remained similar across the Avoidance, Preference, and Neutral Accounts. What changed was the slope of translation equivalent learning: the slopes were the shallowest under the Avoidance Account where BIAS = 0.5, whereas the slopes were steepest under the Preference Account where BIAS = 1.5. With this, we further outline Prediction Set 3:

@@

Ethics approval was obtained by the Human Research Ethics Board of Concordia University (Certification Number 10000439) and informed consent was obtained from the children's parents.

@@

Archival data from 200 bilingual children acquiring English and French (age range: 18.38 - 33.50 months; 94 girls and 106 boys) who participated in prior studies at Concordia Infant Research Lab were included in the present study, drawn from the same set of participants as Gonzalez-Barrero et al.~(2020). Data collection was conducted in Montréal, Québec, Canada. Montréal is a multicultural city where both English and French are widely used in society. Some children took part in more than one in-lab study (n = 28); thus, they contributed data at more than one time point. This resulted in a larger number of datapoints relative to the number of unique participants. The total number of data points included in the analyses was 229 (i.e., 229 English and 229 French CDI questionnaires). Participants were recruited through government birth lists, online ads, daycares, and infant-parent group activities (e.g., children's library activities). Inclusion criteria were the following: full term-pregnancy ( i.e., \textgreater{} 36 weeks of gestation), normal birth weight (\textgreater{} 2500 grams), and absence of major medical conditions (i.e., meningitis). Only children who had complete data in both CDI forms (i.e., English and French) were retained for analysis. Bilingual children were defined as those exposed at least 25\% of the time over the course of their lives globally to both English and French and with less than 10\% of exposure to a third language. For children who participated more than once, their language exposure followed such criteria for all visits. Following the approach in Study 1, children's dominant language was deemed to be the language in which the child produced a greater number of words; vocabulary balance was then determined based on the proportion of words produced in the non-dominant language relative to the total words produced across both languages using the same formula as in Study 1: NONDOM/(DOM+NONDOM). Within the 229 data points, 59.80\% of children were English-dominant and 40.20\% were French-dominant. Children's demographic characteristics including age, maternal education, and language exposure, are presented in Table 4.

@@

Translation equivalents (TE) were determined in the same manner as Gonzalez-Barrero et al.~(2020) by three proficient bilingual French--English adults who carefully examined each language version of the CDI. Word pairs that made reference to the same concept (e.g., English ``apple'' and French \emph{``pomme''}) were considered to be translation equivalents. In cases of disagreement, a discussion of the likely uses of the word in question by children (rather than potential adult uses of the word) was conducted and then a decision was made. Words that had similar phonetic realizations (e.g., English ``alligator'' and French \emph{``alligator''}) were also considered translation equivalents. Most of the items on both vocabulary checklists had an equivalent word in the other language, which resulted in a total of 611 translation equivalents. A full list of translation equivalents is available at {[}\url{https://osf.io/2t5kw/}{]}.

@@

Data analyses were conducted using R (Version 4.0.2, 2020). Analysis scripts and the data set used in the present study are available at {[}\url{https://osf.io/2t5kw/}{]}. We first present descriptive measures of vocabulary, and then tests of the three sets of predictions generated in Study 1.

@@

Children produced an average of 67.70 translation equivalents (TE; SD = 85.10, range = 1 - 409). The remainder of words were singlets, and children produced many more singlets in their dominant language (DOM-SINGLET; M = 138.40, SD = 124.40, range = 2 - 523) than in their non-dominant language (NONDOM-SINGLET; M = 21.20, SD = 20.10, range = 0 - 94), \emph{t}(228) = 13.89, \(p < .001\), \(d = 0.92\). On average, children's concept vocabulary size was 227.30 (CONCEPT; SD = 181.30, range = 4 - 695).

@@

Note that in this paper, we defined BALANCE in terms of relative vocabulary in each language, which numerous studies have found is related to children's relative input in each language (e.g., David \& Wei, 2008; Pearson et al., 1997; Hurtado et al., 2013). We therefore investigated the relationship between vocabulary balance and the percent of exposure bilingual children received in their non-dominant language. For most children, the language in which they produced the most words was also the language that they heard most often (181 children, 79\%), although this was not the case for a minority of children (48 children, 21\%). The correlation between BALANCE and the percentage of exposure to the non-dominant language was moderate, \emph{r}(227) = 0.45, \(p < .001\) (see also Figure 4). The imperfect correlation between these two measures could result from measurement error alone (see Byers-Heinlein et al., 2021, for a discussion of attenuation of correlation due to measurement error in developmental research) or could also be because vocabulary size in each language is determined by factors beyond parent-reported proportion of input in each language, for example the quality of that input (for a discussion of input quality for bilinguals, see Unsworth, 2016).

@@

Prediction Set 2 pertained to expected patterns of acquisition of translation equivalents and singlets for children of different developmental levels. In our data set, developmental level was approximated by children's age. Figure 2 Panel B shows the number of concept vocabulary (CONCEPT), translation equivalents (TE), singlets in the dominant language (DOM-SINGLET) and in the non-dominant language (NONDOM-SINGLET) of the bilingual children as a function of different ages (a proxy for developmental level), used to estimate the number of LEARNABLE words.

@@

Prediction 2c was that both older children and those with the least balanced vocabularies (BALANCE) would produce more dominant-language singlets (DOM-SINGLET). This pattern was confirmed by the results from our dataset, with a positive correlation between dominant-language singlets (DOM-SINGLET) and age (which determined LEARNABLE), \emph{r}(227) = 0.65, \(p < .001\), and a negative correlation between BALANCE and dominant-language singlets (DOM-SINGLET), \emph{r}(227) = -0.58, \(p < .001\). As illustrated in Figure 2 Panel 2C, children with the least balanced vocabulary of 0.1 produced the most singlets in their dominant language (mean = 294), whereas those with the most balanced vocabulary of 0.5 produced the fewest (mean = 66).

Prediction 2d was that older children and those with the most balanced vocabularies (BALANCE) would produce more singlets in their non-dominant language. This pattern was also observed in our dataset, with a positive correlation between the number of non-dominant singlets (NONDOM-SINGLET) and age (which determined LEARNABLE), \emph{r}(227) = 0.19, \(p = .005\), and a positive correlation between BALANCE and the number of non-dominant singlets (NONDOM-SINGLET), \emph{r}(227) = 0.63, \(p < .001\). As shown in Figure 2 Panel 2D, we observed that children produced very few singlets in their non-dominant language, although the most balanced children with BALANCE of 0.5 produced the most singlets in their non-dominant language (mean = 39) and the least balanced children with BALANCE of 0.1 produced the fewest (mean = 4).

@@

First, we will walk through the parameters in this model. The size of dominant vocabulary (DOM) and size of non-dominant vocabulary (NONDOM) were taken to be the number of words produced by individual children observed in the vocabulary data. Again, the number of learnable words (LEARNABLE) was determined by averaging English and French productive CDI vocabulary at the 90th percentile at different ages (percentile information obtained from Wordbank version 0.3.1; Frank et al., 2016). Table 6 lists the denominators derived from averaging these percentiles, or LEARNABLE, at different ages. For example, for an 18 month-old infant, the denominator was 240 words which was calculated by averaging the 259 English words and 220 French words, based on what 18-month-old children would typically produce at the 90th percentile. For children who were between 31 to 33 months in our dataset, the 90th percentile of 30-month-old children was used since 90th percentile information was available only up to 30 months.

@@

\caption{\label{tab:table6}Table 6. The number of total English and French productive CDI vocabulary at the 90th percentile at different ages, and the average between the two which serves as the denominator in our computation model. Values are rounded to the nearest integer.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{>{\centering\arraybackslash}p{75px}>{\centering\arraybackslash}p{125px}>{\centering\arraybackslash}p{125px}>{\centering\arraybackslash}p{125px}}
\toprule
Age (months) & Number of English words produced at 90th percentile & Number of French words produced at 90th percentile & Average (LEARNABLE)\\
\midrule
18 & 259 & 220 & 240\\
19 & 321 & 274 & 298\\
20 & 378 & 325 & 352\\
21 & 430 & 372 & 401\\
22 & 476 & 416 & 446\\
\addlinespace
23 & 517 & 456 & 486\\
24 & 553 & 492 & 522\\
25 & 583 & 525 & 554\\
26 & 610 & 554 & 582\\
27 & 633 & 580 & 606\\
\addlinespace
28 & 653 & 602 & 627\\
29 & 668 & 620 & 644\\
30 - 33 & 681 & 635 & 658\\
\bottomrule
\end{tabular}
\end{table}

Furthermore, the intercept of the linear regression model was set at 0 since no translation equivalents are expected to be produced if a child does not know any dominant or non-dominant vocabulary (i.e., when the predictor variables are 0). To reproduce the Bilingual Vocabulary Model's formula TE = DOM×NONDOM/LEARNABLE, an interaction between dominant and non-dominant vocabulary was entered in the model, but main effects were not included (denoted in R by using a colon rather than an asterisk between the interacting predictors). Therefore, our final linear regression model equation was:

@@

\noindent (In R language, the model formula was rearranged and entered as:

\emph{ObservedTE * 90th percentile of CDI items \textasciitilde{} 0 + Dominant vocabulary:Non-dominant vocabulary})

\noindent With the observed number of translation equivalents as the dependent variable, the regression coefficient estimated by the model would indicate how the BIAS parameter was related to the empirical vocabulary data, which would then indicate whether bilingual children were biased towards or against learning translation equivalents. If the coefficient is close to 1, then there is no bias and translation equivalents are learned equally to other words (i.e., the Neutral Account). Otherwise, a coefficient less than 1 represents a bias against learning translation equivalents where translation equivalents are less easily learned (i.e., the Avoidance Account), and a coefficient greater than 1 represents a bias towards learning translation equivalents where translation equivalents are more easily learned (i.e., the Preference Account).

Our model showed an excellent model fit of \(R^2\) = 0.96, indicating that our model explained 96\% of the variance in bilinguals' translation equivalent knowledge. The linear regression model estimated a BIAS coefficient of 1.04, \(p < .001\). This value is close to 1, suggesting that our data are consistent with the account whereby translation equivalents are learned equivalently to other words\footnote{We also explored how these results would vary if we set the number of LEARNABLE words at the 75th percentile, rather than at the 90th percentile. Overall, model fit was slightly reduced compared to our main model, although fit was still good, \(R^2\) = 0.95, \(p < .001\). The estimated BIAS parameter was lower, at 0.95, which is nonetheless still quite close to 1, and thus consistent with the Neutral Account. This change in the BIAS parameter is mathematically expected due to the reciprocal nature of LEARNABLE and BIAS, an issue we will return to in the discussion.}.

To illustrate the close fit between the Neutral Account and our data, we used the Bilingual Vocabulary Model formula TE = 1 × (DOM×NONDOM/LEARNABLE) to estimate each child's expected translation equivalent knowledge (setting BIAS = 1; also setting the intercept at 0 to replicate our linear regression model), which is plotted against our observed data in Figure 5. Expected and observed translation equivalents were closely aligned with the Neutral Account of the Bilingual Vocabulary Model (i.e., BIAS = 1), suggesting that the Neutral Account provides a parsimonious explanation for bilinguals' translation equivalent knowledge. This provides evidence for the notion that translation equivalents are neither harder nor easier to learn than singlets in bilingual vocabulary learning. Note that visual inspection suggested that there could be some possible outliers. Cook's distance was estimated for our linear regression model listed above and identified two data points with a cook's distance over 0.4. After removing those two data points, the linear regression model returned a coefficient of 1.07, \(p < .001\), with \(R^2\) = 0.96. As the model fit was similar to the model without eliminating the two outlier data points, we proceeded with the full data set keeping the two potential outlier data points.

@@

Despite the good overall fit to the data, a close examination of Figure 5 suggested that the model might less closely fit the data of children with smaller vocabulary sizes. Figure 7 displays the model fit separately for children with a word vocabulary (WORD) less than 300 words and those with a word vocabulary (WORD) of 300 or greater. Based on visual inspection, the slope of translation equivalent learning appeared steeper for children with less than 300 total vocabulary, suggesting that translation equivalents are more easily learned (i.e., BIAS \textgreater{} 1); whereas the slope of translation equivalent learning appeared to align with the Neutral Account (i.e., BIAS = 1) for children with more than 300 total vocabulary. To further explore this pattern, we ran the same linear regression twice, separately for children with less than 300 total vocabulary and for those with more than 300 total vocabulary. The model for those with larger total word vocabulary (WORD) returned a coefficient of BIAS = 1.04, \(p < .001\), whereas the model for those with less than 300 total word vocabulary (WORD) returned a coefficient of BIAS = 2.21, \(p < .001\). Both models fit well, although a somewhat better fit was obtained for children with larger vocabulary size (\(R^2\) = 0.96) than children with smaller vocabulary size (\(R^2\) = 0.88). Overall, this analysis suggests that translation equivalent learning for children with larger vocabularies corresponds best to the Neutral Account, but translation equivalent learning for children with smaller vocabularies corresponds best to the Preference Account.

@@

Another possible explanation is that the nature of bilingual input changes as children become more advanced word learners. Some recent research has suggested that bilingual parents sometimes code-switch to use a word that they know to be in their child's vocabulary (Kremin et al., 2021; Nicoladis \& Secco, 2000). For example, a caregiver may choose to say to their English--French bilingual child ``Can you give me the \emph{livre}?'' if they know their child understands the French word \emph{``livre''} but not the English equivalent ``book.'' This may provide fewer opportunities for children to learn translation equivalents, since they would be less exposed to the unfamiliar translation equivalents. However, this observation would predict that young bilinguals would know fewer translation equivalents as a proportion of their vocabularies than older bilinguals, which was opposite to what we observed. Thus, changes in bilingual input also do not provide an adequate explanation for our results of a qualitative change in translation equivalent learning. Overall, more research will be needed to understand why translation equivalents appear to be over-represented in younger bilinguals' vocabularies.

Finally, the developmental pattern we observed could be due to differences in how age and developmental level are related across infancy, rather than to word learning biases. Indeed, even controlling for age, vocabulary size is linked to multiple factors, including cognitive abilities (e.g., Reuter et al., 2018), speech perception skills (e.g., Cristia et al., 2014), and amount of language experience (e.g., Weisleder \& Fernald, 2013). Our model used age as a proxy for developmental level, assuming that all children of a particular age could potentially learn the same number of words as children at the 90th vocabulary percentile. However, there could be much more variability in the developmental level of younger infants than of older infants. In this case, the 90th percentile might be a reasonable estimate for the number of learnable words for most older infants, but an overestimate of the number of learnable words for all but the most cognitively advanced younger infants. Given the importance of the learnability parameter in our models, future research will be needed to more precisely quantify the number of words that are learnable by particular children at particular ages.

@@

>>>>

\abstract{
The acquisition of translation equivalents is often considered a special component of bilingual children's vocabulary development, as bilinguals have to learn words that share the same meaning across their two languages. This study examined three contrasting accounts for bilingual children's acquisition of translation equivalents relative to words that are first labels for a referent: the Avoidance Account whereby translation equivalents are harder to learn, the Preference Account whereby translation equivalents are easier to learn, and the Neutral Account whereby translation equivalents are similar to learn. To adjudicate between these accounts, Study 1 explored patterns of translation equivalent learning under a novel computational model --- the Bilingual Vocabulary Model --- which quantifies translation equivalent knowledge as a function of the probability of learning words in each language, and includes a bias parameter that varies the difficulty of learning translation equivalents according to each account. Study 2 tested model-derived predictions against vocabulary data from 200 French--English bilingual children aged 18--33 months. Results showed a close match between the model predictions and bilingual children's patterns of translation equivalent learning. At smaller vocabulary sizes, data matched the Preference Account, while at larger vocabulary sizes they matched the Neutral Account. Our findings show that patterns of translation equivalent learning emerge predictably from the word learning process, and potentially reveal a qualitative shift in translation equivalent learning as bilingual children develop and learn more words.
}

@@

\begin{verbatim}
## Warning: Removed 20 rows containing non-finite values (stat_smooth).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 140 rows containing missing values (geom_smooth).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 20 rows containing non-finite values (stat_smooth).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 140 rows containing missing values (geom_smooth).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 3 rows containing non-finite values (stat_smooth).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 25 rows containing missing values (geom_smooth).
\end{verbatim}

@@

\caption{Number of translation equivalents (TE) across different levels of vocabulary balance (BALANCE) in relation to word vocabulary (WORD; Panel A), dominant vocabulary size (DOM; Panel B), and non-dominant vocabulary size (NONDOM; Panel C). Row 1 plots the simulated data in Study 1 while holding the number of learnable words (LEARNABLE) constant at 600 and BIAS constant at 1. Row 2 plots the observed vocabulary data in Study 2.}\label{fig:fig1}
\end{figure}

@@

\begin{verbatim}
## Warning: Removed 39 rows containing non-finite values (stat_smooth).
\end{verbatim}

\begin{verbatim}
## Warning: Removed 1 rows containing missing values (geom_smooth).
\end{verbatim}

@@

Translation equivalent (TE) knowledge was first simulated across different developmental levels (as indicated by number of LEARNABLE words = 150, 300, 450, 600), in conjunction with a wide range of values for DOM and NONDOM. Again, BALANCE and word vocabulary (WORD) were calculated based on the values of DOM and NONDOM. The final simulated data set contained 166 data points (see Table 2 for a summary of the parameters used). Three scenarios of translation equivalent learning (TE) were then generated using the formula TE = BIAS × DOM×NONDOM/LEARNABLE. To illustrate the Avoidance Account, BIAS was set at .5 (i.e., TEs are 50\% less likely to be learned than singlets). To illustrate the Neutral Account, BIAS was set at 1 (i.e., TEs are equally learnable as singlets). Finally, to illustrate the Preference Account, BIAS was set at 1.5 (i.e., TE are 50\% more likely to be learned than singlets). In Figure 3, we illustrate the three different scenarios of simulated translation equivalent (TE) knowledge. Again, we observed a pattern consistent with prediction 1a where, in all cases, simulated children with more balanced vocabularies (BALANCE) produced more translation equivalents (TE). Thus, overall relationships between BALANCE and TE remained similar across the Avoidance, Preference, and Neutral Accounts. What changed was the slope of translation equivalent learning: the slopes were the shallowest under the Avoidance Account where BIAS = 0.5, whereas the slopes were steepest under the Preference Account where BIAS = 1.5. With this, we further outline Prediction Set 3:

@@

Ethics approval was obtained by the Human Research Ethics Board of Concordia University (Certificate Number 10000439) and informed consent was obtained from the children's parents.

@@

Archival data from 200 bilingual children acquiring English and French (age range: 18.40 - 33.50 months; 94 girls and 106 boys) who participated in prior studies at Concordia Infant Research Lab were included in the present study, drawn from the same set of participants as Gonzalez-Barrero et al.~(2020). Data collection was conducted in Montréal, Québec, Canada. Montréal is a multicultural city where both English and French are widely used in society. Some children took part in more than one in-lab study (n = 28); thus, they contributed data at more than one time point. This resulted in a larger number of datapoints relative to the number of unique participants. The total number of data points included in the analyses was 229 (i.e., 229 English and 229 French CDI questionnaires). Participants were recruited through government birth lists, online ads, daycares, and infant-parent group activities (e.g., children's library activities). Inclusion criteria were the following: full term-pregnancy ( i.e., \textgreater{} 36 weeks of gestation), normal birth weight (\textgreater{} 2500 grams), and absence of major medical conditions (i.e., meningitis). Only children who had complete data in both CDI forms (i.e., English and French) were retained for analysis. Bilingual children were defined as those exposed at least 25\% of the time over the course of their lives globally to both English and French and with less than 10\% of exposure to a third language. For children who participated more than once, their language exposure followed such criteria for all visits. Following the approach in Study 1, children's dominant language was deemed to be the language in which the child produced a greater number of words; vocabulary balance was then determined based on the proportion of words produced in the non-dominant language relative to the total words produced across both languages using the same formula as in Study 1: NONDOM/(DOM+NONDOM). Within the 229 data points, 59.80\% of children were English-dominant and 40.20\% were French-dominant. Children's demographic characteristics including age, maternal education, and language exposure, are presented in Table 4.

@@

Translation equivalents (TE) were determined in the same manner as Gonzalez-Barrero et al.~(2020) by three proficient bilingual French--English adults who carefully examined each language version of the CDI. Word pairs that made reference to the same concept (e.g., English ``apple'' and French \emph{``pomme''}) were considered to be translation equivalents. In cases of disagreement, a discussion of the likely uses of the word in question by children (rather than potential adult uses of the word) was conducted and then a decision was made. Words that had similar phonetic realizations (e.g., English ``alligator'' and French \emph{``alligator''}) were also considered translation equivalents. Most of the items on both vocabulary checklists had an equivalent word in the other language, which resulted in a total of 611 translation equivalents. A full list of translation equivalents is available at \url{https://osf.io/2t5kw/}.

@@

Data analyses were conducted using R (Version 4.0.2, 2020). Analysis scripts and the data set used in the present study are available at \url{https://osf.io/2t5kw/}. We first present descriptive measures of vocabulary, and then tests of the three sets of predictions generated in Study 1.

@@

Children produced an average of 67.70 translation equivalents (TE; SD = 85.10, range = 1 - 409). The remainder of words were singlets, and children produced many more singlets in their dominant language (DOM-SINGLET; M = 138.40, SD = 124.40, range = 2 - 523) than in their non-dominant language (NONDOM-SINGLET; M = 21.20, SD = 20.10, range = 0 - 94), \emph{t}(228) = 13.89, \(p < .001\), \(d = 0.92\). On average, children's concept vocabulary size was 227.30 words (CONCEPT; SD = 181.30, range = 4 - 695).

@@

Note that in this paper, we defined BALANCE in terms of relative vocabulary in each language, which numerous studies have found is related to children's relative input in each language (e.g., David \& Wei, 2008; Pearson et al., 1997; Hurtado et al., 2013). We therefore investigated the relationship between vocabulary balance and the percent of exposure bilingual children received in their non-dominant language. For most children, the language in which they produced the most words was also the language that they heard most often (181 children, 79\%), although this was not the case for a minority of children (48 children, 21\%). The correlation between BALANCE and the percentage of exposure to the non-dominant language was moderate, \emph{r}(227) = 0.45, \(p < .001\) (see also Figure 4). The imperfect correlation between these two measures could result from measurement error (see Byers-Heinlein et al., 2021, for a discussion of attenuation of correlation due to measurement error in developmental research) or could also be because vocabulary size in each language is determined by factors beyond parent-reported proportion of input in each language, for example the quality of that input (for a discussion of input quality for bilinguals, see Unsworth, 2016).

@@

Prediction Set 2 pertained to expected patterns of acquisition of translation equivalents and singlets for children of different developmental levels. In our data set, developmental level was approximated by children's age. Figure 2 Panel B shows the number of concept vocabulary (CONCEPT), translation equivalents (TE), singlets in the dominant language (DOM-SINGLET) and in the non-dominant language (NONDOM-SINGLET) of the bilingual children as a function of different ages (our proxy for developmental level), used to estimate the number of LEARNABLE words.

@@

Prediction 2c was that both older children and those with the least balanced vocabularies (BALANCE) would produce more dominant-language singlets (DOM-SINGLET). This pattern was confirmed by the results from our dataset, with a positive correlation between dominant-language singlets (DOM-SINGLET) and age (which determined LEARNABLE), \emph{r}(227) = 0.65, \(p < .001\), and a negative correlation between BALANCE and dominant-language singlets (DOM-SINGLET), \emph{r}(227) = -0.58, \(p < .001\). This is illustrated in Figure 2 Panel 2C.

Prediction 2d was that older children and those with the most balanced vocabularies (BALANCE) would produce more singlets in their non-dominant language. This pattern was also observed in our dataset, with a positive correlation between the number of non-dominant singlets (NONDOM-SINGLET) and age (which determined LEARNABLE), \emph{r}(227) = 0.19, \(p = .005\), and a positive correlation between BALANCE and the number of non-dominant singlets (NONDOM-SINGLET), \emph{r}(227) = 0.63, \(p < .001\). This is illustrated in Figure 2 Panel 2D.

@@

First, we will walk through the parameters in this model. The size of dominant vocabulary (DOM) and size of non-dominant vocabulary (NONDOM) were taken to be the number of words produced by individual children observed in the vocabulary data. Again, the number of learnable words (LEARNABLE) was determined by averaging English and French productive CDI vocabulary at the 90th percentile at different ages (percentile information obtained from Wordbank version 0.3.1; Frank et al., 2016). Table 6 lists the denominators derived from averaging these percentiles, or LEARNABLE, at different ages. For example, for an 18 month-old infant, the denominator was 240 words which was calculated by averaging the 259 English words and 220 French words, based on what 18-month-old children would typically produce at the 90th percentile. For children who were between 31 to 33 months in our dataset, the 90th percentile of 30-month-old children was used since 90th percentile information was available only up to 30 months. Finally, the intercept of the linear regression model was set at 0 since no translation equivalents are expected to be produced if a child does not know any dominant or non-dominant vocabulary (i.e., when the predictor variables are 0).

@@

\caption{\label{tab:table6}Table 6. The number of total English and French productive CDI vocabulary at the 90th percentile at different ages according to WordBank, and the average between the two which serves as the denominator in our computation model. Values are rounded to the nearest integer.}
\centering
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{>{\centering\arraybackslash}p{75px}>{\centering\arraybackslash}p{125px}>{\centering\arraybackslash}p{125px}>{\centering\arraybackslash}p{125px}}
\toprule
Age (months) & Number of English words produced at 90th percentile & Number of French words produced at 90th percentile & Average (LEARNABLE)\\
\midrule
18 & 259 & 220 & 240\\
19 & 321 & 274 & 298\\
20 & 378 & 325 & 352\\
21 & 430 & 372 & 401\\
22 & 476 & 416 & 446\\
\addlinespace
23 & 517 & 456 & 486\\
24 & 553 & 492 & 522\\
25 & 583 & 525 & 554\\
26 & 610 & 554 & 582\\
27 & 633 & 580 & 606\\
\addlinespace
28 & 653 & 602 & 627\\
29 & 668 & 620 & 644\\
30 - 33 & 681 & 635 & 658\\
\bottomrule
\end{tabular}
\end{table}

To reproduce the Bilingual Vocabulary Model's formula TE = DOM×NONDOM/LEARNABLE, an interaction between dominant and non-dominant vocabulary was entered in the model, but main effects were not included (denoted in R by using a colon rather than an asterisk between the interacting predictors). Therefore, our final linear regression model equation was\footnote{Note that for the purposes of running the model in R, the formula was rearranged and entered as \emph{ObservedTE * 90th percentile of CDI items \textasciitilde{} 0 + Dominant vocabulary:Non-dominant vocabulary}.}:

@@

With the observed number of translation equivalents as the dependent variable, the regression coefficient estimated by the model indicates how the BIAS parameter is related to the empirical vocabulary data, which in turn indicates whether bilingual children were biased towards or against learning translation equivalents. If the coefficient is close to 1, then there is no bias and translation equivalents are learned equally to other words (i.e., the Neutral Account). Otherwise, a coefficient less than 1 indicates a bias against learning translation equivalents where translation equivalents are less easily learned (i.e., the Avoidance Account), and a coefficient greater than 1 indicates a bias towards learning translation equivalents where translation equivalents are more easily learned (i.e., the Preference Account).

Our model showed an excellent model fit of \(R^2\) = 0.96, indicating that our model explained 96\% of the variance in bilinguals' translation equivalent knowledge. BIAS coefficient was estimated at 1.04. This value is close to 1, suggesting that our data are consistent with the account whereby translation equivalents are learned equivalently to other words\footnote{We also explored how these results would vary if we set the number of LEARNABLE words at the 75th percentile, rather than at the 90th percentile. Overall, model fit was slightly reduced compared to our main model, although fit was still good, \(R^2\) = 0.95. The estimated BIAS parameter was lower, at 0.95, which is nonetheless still quite close to 1, and thus consistent with the Neutral Account. This change in the BIAS parameter is mathematically expected due to the reciprocal nature of LEARNABLE and BIAS, an issue we will return to in the discussion.}.

To illustrate the close fit between the Neutral Account and our data, we used the Bilingual Vocabulary Model formula TE = 1 × (DOM×NONDOM/LEARNABLE) to estimate each child's expected translation equivalent knowledge (setting BIAS = 1; also setting the intercept at 0 to replicate our linear regression model), which is plotted against our observed data in Figure 5. Expected and observed translation equivalents were closely aligned with the Neutral Account of the Bilingual Vocabulary Model (i.e., BIAS = 1), suggesting that the Neutral Account provides a parsimonious explanation for bilinguals' translation equivalent knowledge. This provides evidence for the notion that translation equivalents are neither harder nor easier to learn than singlets in bilingual vocabulary learning. Note that visual inspection suggested that there could be some possible outliers. Cook's distance was estimated for our linear regression model listed above and identified two data points with a cook's distance over 0.4. After removing those two data points, the linear regression model returned a coefficient of 1.07, with \(R^2\) = 0.96. As the model fit was similar to the model without eliminating the two outlier data points, we proceeded with the full data set keeping the two potential outlier data points.

@@

Despite the good overall fit to the data, a close examination of Figure 5 suggested that the model might less closely fit the data of children with smaller vocabulary sizes. Figure 7 displays the model fit separately for children with a word vocabulary (WORD) less than 300 words and those with a word vocabulary (WORD) of 300 or greater. Based on visual inspection, the slope of translation equivalent learning appeared steeper for children with less than 300 total vocabulary, suggesting that translation equivalents are more easily learned (i.e., BIAS \textgreater{} 1); whereas the slope of translation equivalent learning appeared to align with the Neutral Account (i.e., BIAS = 1) for children with more than 300 total vocabulary. To further explore this pattern, we ran the same linear regression twice, separately for children with less than 300 total vocabulary and for those with more than 300 total vocabulary. The model for those with larger total word vocabulary (WORD) returned a coefficient of BIAS = 1.04, whereas the model for those with less than 300 total word vocabulary (WORD) returned a coefficient of BIAS = 2.21. Both models fit well, although a somewhat better fit was obtained for children with larger vocabulary size (\(R^2\) = 0.96) than children with smaller vocabulary size (\(R^2\) = 0.88). Overall, this analysis suggests that translation equivalent learning for children with larger vocabularies corresponds best to the Neutral Account, but translation equivalent learning for children with smaller vocabularies corresponds best to the Preference Account.

@@

Another possible explanation is that the nature of bilingual input changes as children become more advanced word learners. Some recent research has suggested that bilingual parents sometimes code-switch to use a word that they know to be in their child's vocabulary (Kremin et al., 2021; Nicoladis \& Secco, 2000). For example, a caregiver may choose to say to their English--French bilingual child ``Can you give me the \emph{livre}?'' if they know their child understands the French word ``\emph{livre}'' but not the English equivalent ``book.'' This may provide fewer opportunities for children to learn translation equivalents, since they would be less exposed to the unfamiliar translation equivalents. However, this observation would predict that young bilinguals would know fewer translation equivalents as a proportion of their vocabularies than older bilinguals, which was opposite to what we observed. Thus, changes in bilingual input also do not provide an adequate explanation for our results of a qualitative change in translation equivalent learning.

Finally, the developmental pattern we observed could be due to differences in how age and developmental level are related across infancy, rather than to changes in input patterns or word learning biases. Indeed, even controlling for age, vocabulary size is linked to multiple factors, including cognitive abilities (e.g., Reuter et al., 2018), speech perception skills (e.g., Cristia et al., 2014), and amount of language experience (e.g., Weisleder \& Fernald, 2013). Our model used age as a proxy for developmental level, assuming that all children of a particular age could potentially learn the same number of words as children at the 90th vocabulary percentile. However, there could be much more variability in the developmental level of younger infants than of older infants. In this case, the 90th percentile might be a reasonable estimate for the number of learnable words for most older infants, but an overestimate of the number of learnable words for all but the most cognitively advanced younger infants. Given the importance of the learnability parameter in our models, future research will be needed to more precisely quantify the number of words that are learnable by particular children at particular ages.

@@

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

\noindent We are grateful to all the families who participated in this research. This work was funded by a grant from the Natural Sciences and Engineering Research Council of Canada (NSERC 2011-402470, 2018-04390), a National Institutes of Health grant (NIH R03HD079779), and a Fonds de Recherche du Québec-Société et Culture grant (FQRSC 2012-NP-145009) to Byers-Heinlein, by a FQRSC (2018-B3-205717) postdoctoral fellowship to Gonzalez-Barrero, and by a FQRSC graduate fellowship and Concordia University graduate fellowship to Schott. Byers-Heinlen holds the Concordia University Research Chair in Bilingualism and Open Science. This work was presented at the 2021 Boston University Conference on Language Development. We thank the members of the Concordia Infant Research Lab for their comments on earlier versions of this paper, and Roman Feiman for thoughtful discussion that led to new insights.

~
~

\hypertarget{author-contributions}{%
\section{Author Contributions}\label{author-contributions}}
